Stage-Splitting a Modal Language

### 2014-08-05 paper/semantics.tex
moving around text in semantics section

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code. 
Therefore, a suitably staged dynamic semantics for \lang\ must evaluate
\emph{all} the stage \bbone\ subexpressions of a term before its stage
\bbtwo\ subexpressions.

@@

To gain intuition about the challenges of ensuring staged evaluation of \lang, consider the following
example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {1+2}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
then obtain \verb|7| as the result of projection.

@@

\subsection{Dynamics}

@@

To fully evaluate an expression in \lang, we first evaluate all of its stage
\bbone\ subexpressions as described above. This results in a stage \bbtwo\
expression with no stage \bbone\ content; we say that it is a term in a
monostaged%
\footnote{In general, \lang could have more than two stages, in which case the
expression would be in a language with one fewer stage.}
language called \langTwo. Ordinary \langTwo\ evaluation, implemented by the
$\tworedsym$ judgment, completes the evaluation of this term.

>>>>

Although the code for {\tt fexp} appears very similar to the unstaged mathematical
definition above, it is in fact a staged program: the {\tt if} predicates and exponent
decomposition are all stage-\bbone\ terms, since they occur within $\prev$ blocks.
%\needsfix{Essentially, the type system is sufficient to prove the observation that all of
%the decomposition of the exponent can be done at the stage \bbone\ without
%requiring disruption of the elegant functional structure of the code.}
%
% NOTE(kayvonf): removing the following comments about binding time analysis from the text for now to improve clarity.  They will certainly be addressed in some prior section in the paper.
%
%We note that our work does not consider the problem of {\em binding time analysis}---transforming an unstaged program into a staged \lang\ program by inserting appropriate $\next$, $\prev$, and $\pause$ constructs. We assume that binding time analysis is performed manually by a programmer (as done above) or via automatic analysis of unstaged code to arrive at a valid \lang\ program used as input for subsequent splitting.
%
A key attribute of this example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside
stage \bbtwo\ ones (such as the recurive call to {\tt fexp} above), this strategy conflicts with the precept of staged execution: that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage \bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage \bbone\ content; we say this is a term in a
monostaged%
\footnote{In general, \lang could have more than two stages, in which case the
expression would be in a language with one fewer stage.}
language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo.

@@

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the following example:
\begin{lstlisting}
#2 (next {1+2}, 3+4)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. A conventional
call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {1+2}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|3|), evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {1+2}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {1+2}, 7)
\end{lstlisting}
then obtain \verb|7| as the result of projection.

@@

\subsection{Dynamics}
\label{ssec:dynamics}

@@
